
{
\setlength{\bibhang}{2em}
\def\newblock{\hskip .11em plus .33em minus .07em}

\hspace{\parindent} Keuchel, S. and Jeuring, J.~T. (2012).
\newblock Generic conversions of abstract syntax representations.
\newblock In {\em Proceedings of the 8th ACM SIGPLAN workshop on Generic
  programming, \textup{WGP '12}}, pages 57--68. ACM.
\newblock Copenhagen, Denmark, September 12, 2012.
\vspace{3mm}


Delaware, B., Keuchel, S., Schrijvers, T., and Oliveira,
B. C. d.~S. (2013).
\newblock Modular Monadic Meta-Theory.
\newblock In {\em Proceedings of the 18th ACM SIGPLAN international
  conference on Functional programming}, ICFP '13, pages 319-330. ACM.
\newblock Boston, Massachusetts, USA, September 25--27, 2013.
\vspace{3mm}


Keuchel, S. and Schrijvers, T. (2013).
\newblock Generic Datatypes \`a la Carte.
\newblock In {\em Proceedings of the 9th ACM SIGPLAN workshop on Generic
  programming}, WGP ’13, pages 13-24. ACM.
\newblock Boston, Massachusetts, USA, September 28, 2013.
\vspace{3mm}


Keuchel, S., Weirich, S., and Schrijvers, T. (2016).
\newblock {Needle {\&} Knot: Binder Boilerplate Tied Up}.
\newblock In Thiemann, P., editor, {\em Proceedings of the 25th European
  Symposium on Programming, \textup{ESOP'16}}, volume 9632 of {\em Lecture
  Notes in Computer Science}, pages 419--445. Springer.
\newblock Eindhoven, The Netherlands, April 2--8, 2016.
\vspace{3mm}


Devriese, D., Patrignani, M., Piessens, F., and Keuchel, S. (2017).
\newblock {Modular, Fully-abstract Compilation by Approximate
  Back-translation}.
\newblock {\em {Logical Methods in Computer Science}}, {Volume 13, Issue 4}.

}
